Type Refinements for the Working Class

### 2015-11-26 paper.tex
Move some macros to a new package
First cut definition of functional equality; fix bugs

\usepackage{notation/modes}
\usepackage{notation/judgments}
\usepackage{notation/common}
\usepackage{notation/abts}

@@

\newcommand\MapsTo[2]{\IMode{#1}\mapsto\OMode{#2}}

@@

\newcommand\IsSubsetEq[2]{\IMode{#1}\subseteq\IMode{#2}}

\newcommand\Embed[1]{\left\lfloor#1\right\rfloor}

% parametric refinement
\newcommand\Refines[3]{\IMode{#1}\parallel\IMode{#2}\sqsubset\OMode{#3}}

% global parametric refinement
\newcommand\GloballyRefines[2]{\IMode{#1}\sqsubset\OMode{#2}}

% equality of parametric refinement
\newcommand\EqRefines[4]{\IMode{#1}\parallel\IMode{#2}=\IMode{#3}\sqsubset\OMode{#4}}

% parametric context refinement
\newcommand\RefinesCtx[3]{\IMode{#1}\parallel\IMode{#2}\sqsubset^\star\OMode{#3}}

% functional refinement
\newcommand\FRefines[4]{%
  \IMode{#1}\parallel\IMode{#2}%
  \vDash\IMode{#3}\sqsubset\OMode{#4}%
}

% equality of functional refinement
\newcommand\FEqRefines[5]{%
  \IMode{#1}\parallel\IMode{#2}%
  \vDash\IMode{#3}=\IMode{#4}
  \sqsubset\OMode{#5}%
}

% parametric equality of refinement members
\newcommand\RMemEq[4]{\IMode{#1}\parallel\IMode{#2}=\IMode{#3}\in\IMode{#4}}

% functional equality of refinement members
\newcommand\FRMemEq[5]{\IMode{#1}\parallel\IMode{#2}\vDash\IMode{#3}=\IMode{#4}\in\IMode{#5}}

% parametric subrefinement
\newcommand\IsSubrefinement[4]{%
  \IMode{#1}\parallel\IMode{#2}%
  \subseteq\IMode{#3}\sqsubset\OMode{#4}%
}

@@

\begin{quote}
  To know $\FEqRefines{\Upsilon}{\Psi,x:\chi}{\phi}{\psi}{\tau}$ is to know, for any renaming
  \begin{tikzcd}[cramped]
    \IMode{\Upsilon} \arrow[r,hook,"\IMode{\rho}"] &\IMode{\Upsilon'}
  \end{tikzcd}
  and closed terms $\Member{M,N}{\Exprs_\sigma(\Upsilon')}$, that from
  $\FRMemEq{\Upsilon}{\Psi\rho}{M}{N}{\chi\rho}$, you can conclude
  $\FEqRefines{\Upsilon}{\Psi\rho}{\Subst{M}{x}{\phi\rho}}{\Subst{N}{x}{\psi\rho}}{\tau}$.
  In other words:
  %
  \[
    \infer[\PFunRefSnoc]{
      \FEqRefines{\Upsilon}{\Psi,x:\chi}{\phi}{\psi}{\tau}
    }{
      \begin{array}{l}
        \forall
        \begin{tikzcd}[cramped, ampersand replacement = \&]
          \IMode{\Upsilon} \arrow[r,hook,"\IMode{\rho}"] \&\IMode{\Upsilon'}
        \end{tikzcd}.\
        \forall \Member{M,N}{\Exprs_\sigma(\Upsilon')}.\\
        \left(\FRMemEq{\Upsilon}{\Psi\rho}{M}{N}{\chi\rho}\right)
        \Rightarrow
        \left(\FEqRefines{\Upsilon}{\Psi\rho}{\Subst{M}{x}{\phi\rho}}{\Subst{N}{x}{\psi\rho}}{\tau}\right)
      \end{array}
    }
  \]
\end{quote}

%%
%% \subsection{Functional Equality}
%%
%% Functional equality $\FRMemEq{\Phi}{\Psi}{M}{N}{\phi}$ presupposes
%% $\RefinesCtx{\Phi}{\Upsilon}$, $\RefinesCtx{\Phi}{\Psi}{\Gamma}$,
%% $\FRefines{\Phi}{\Psi}{\phi}{\tau}$, $\IsAbt{\cdot}{\Upsilon}{\Gamma}{M}{\tau}$
%% and $\IsAbt{\cdot}{\Upsilon}{\Gamma}{N}{\tau}$; the judgment is defined by
%% induction on the evidence for the second presupposition
%% $\RefinesCtx{\Phi}{\Psi}{\Gamma}$:
%%
%% \paragraph{Case} $\DefRefinesCtxNil{\Phi}$.
%% \[
%%   \infer{
%%     \FRMemEq{\Phi}{\cdot}{M}{N}{\phi}
%%   }{
%%     \RMemEq{\Phi}{M}{N}{\phi}
%%   }
%% \]
%%
%% \paragraph{Case} $\DefRefinesCtxSnoc{\Phi}{\Psi}{x}{\psi}{\Gamma}{\sigma}$,
%% where $\RefinesCtx{\Phi}{\Upsilon}$.
%% \[
%%   \infer{
%%     \FRMemEq{\Phi}{\Psi,x:\psi}{M}{N}{\phi}
%%   }{
%%     \HPGenJ{\Upsilon}{\Gamma}{\MV{m},\MV{n}:\sigma}{
%%       \HypJ{
%%         \FRMemEq{\Phi}{\Psi}{
%%           \Subst{\MV{m}}{x}{\phi}
%%         }{
%%           \Subst{\MV{n}}{x}{\psi}
%%         }{\phi}
%%       }{
%%         \FRMemEq{\Phi}{\Psi}{\MV{m}}{\MV{n}}{\psi}
%%       }
%%     }
%%   }
%% \]
%%
%% \section{Constructions on Refinements}
%%
%% Refinements can be arranged in a partial order via the
%% $\IsSubrefinement{\Phi}{\phi}{\psi}{\tau}$ judgment, and admit a lattice structure
%% at any sort $\tau$.
%%
%% We have the bottom refinement $\GloballyRefines{\TyBot{\tau}}{\tau}$, which simply
%% contains no elements. Next, the top refinement $\GloballyRefines{\TyTop{\tau}}{\tau}$
%% identifies all terms $M,N$ such that $\IsAbt{\cdot}{\Upsilon}{\cdot}{M}{\tau}$
%% and $\IsAbt{\cdot}{\Upsilon}{\cdot}{N}{\tau}$. The join
%% $\Refines{\Phi}{\phi\TyJoin{\tau}\psi}{\tau}$ of two refinements
%% $\Refines{\Phi}{\phi}{\tau}$ and $\Refines{\Phi}{\psi}{\tau}$ is defined as follows:
%% \begin{gather*}
%%   \infer{
%%     \RMemEq{\Phi}{M}{N}{\phi\TyJoin{\tau}\psi}
%%   }{
%%     \RMemEq{\Phi}{M}{N}{\phi}
%%   }\qquad
%%   \infer{
%%     \RMemEq{\Phi}{M}{N}{\phi\TyJoin{\tau}\psi}
%%   }{
%%     \RMemEq{\Phi}{M}{N}{\psi}
%%   }
%% \end{gather*}
%%
%% Finally the meet $\Refines{\Phi}{\phi\TyMeet{\tau}\psi}{\tau}$ of two refinements
%% $\Refines{\Phi}{\phi}{\tau}$ and $\Refines{\Phi}{\psi}{\tau}$ is defined
%% dually:
%% \[
%%   \infer{
%%     \RMemEq{\Phi}{M}{N}{\phi\TyMeet{\tau}\psi}
%%   }{
%%     \RMemEq{\Phi}{M}{N}{\phi} &
%%     \RMemEq{\Phi}{M}{N}{\psi}
%%   }
%% \]
%%
%% It is trivial to verify that these constructions on refinements give rise to a
%% lattice.
%%
%% \section{Multi-Sorted Computational Type Theory}
%%
%% Martin-L\"of's type theory is a theory of \emph{value types}: to define a type,
%% you specify how to form its equal canonical inhabitants, or values. Then, the
%% inhabitants of the types are explained uniformly by extending this relation
%% over an operational semantics such that two terms are equal members of a
%% type just when they compute to equal values of that type.
%%
%% More generally, one may wish to consider types which include computations in
%% addition to values; an example of such a type would be $\bar{\psi}$ when $\psi$
%% is a type, which would include all the members of $\psi$ as well as a divergent
%% computation. Another example is the type $\TyTop{\tau}$ which identifies all
%% closed terms of sort $\tau$.
%%
%% In order to ensure that such types are sensible, we must formulate a notion of
%% computational equivalence, and then say that a refinement is a \emph{type} just
%% when it respects computational equivalence; then, we are able to define types
%% as PERs on \emph{programs} in addition to values.

>>>>

\usepackage{notation/modes}
\usepackage{notation/judgments}
\usepackage{notation/common}
\usepackage{notation/abts}
\usepackage{refinements}

@@

\newcommand\MapsTo[2]{\IMode{#1}\mapsto\OMode{#2}}
\newcommand\IsSubsetEq[2]{\IMode{#1}\subseteq\IMode{#2}}

@@

@@

\begin{quote}
  To know $\FEqRefines{\Upsilon}{\Psi,x:\chi}{\phi}{\psi}{\tau}$ is to know, for any renaming
  \begin{tikzcd}[cramped]
    \IMode{\Upsilon} \arrow[r,hook,"\IMode{\rho}"] &\IMode{\Upsilon'}
  \end{tikzcd}
  and closed terms $\Member{M_0,M_1}{\Exprs_\sigma(\Upsilon')}$, that from
  $\FRMemEq{\Upsilon'}{\Psi\rho}{M_0}{M_1}{\chi\rho}$, you can conclude
  $\FEqRefines{\Upsilon'}{\Psi\rho}{\Subst{M_0}{x}{\phi\rho}}{\Subst{M_1}{x}{\psi\rho}}{\tau}$.
  In other words:
  %
  \[
    \infer[\PFunRefSnoc]{
      \FEqRefines{\Upsilon}{\Psi,x:\chi}{\phi}{\psi}{\tau}
    }{
      \begin{array}{l}
        \forall
        \begin{tikzcd}[cramped, ampersand replacement = \&]
          \IMode{\Upsilon} \arrow[r,hook,"\IMode{\rho}"] \&\IMode{\Upsilon'}
        \end{tikzcd}.\
        \forall \Member{M_0,M_1}{\Exprs_\sigma(\Upsilon')}.\\
        \left(\FRMemEq{\Upsilon'}{\Psi\rho}{M_0}{M_1}{\chi\rho}\right)
        \Rightarrow
        \left(\FEqRefines{\Upsilon'}{\Psi\rho}{\Subst{M_0}{x}{\phi\rho}}{\Subst{M_1}{x}{\psi\rho}}{\tau}\right)
      \end{array}
    }
  \]
\end{quote}

\subsection{Functional Equality}

Functional equality $\FRMemEq{\Upsilon}{\Psi}{M}{N}{\phi}$ presupposes
$\RefinesCtx{\Upsilon}{\Psi}{\Gamma}$, $\FRefines{\Upsilon}{\Psi}{\phi}{\tau}$,
and $\Member{M,N}{\Exprs_\tau(\Upsilon)}$; this judgment is defined by
induction on the evidence for the first presupposition
$\RefinesCtx{\Upsilon}{\Psi}{\Gamma}$:

\paragraph{Case} $\DefRefinesCtxNil{\Upsilon}$.
\[
  \infer{
    \FRMemEq{\Upsilon}{\cdot}{N_0}{N_1}{\phi}
  }{
    \RMemEq{\Upsilon}{N_0}{N_1}{\phi}
  }
\]

\paragraph{Case} $\DefRefinesCtxSnoc{\Upsilon}{\Psi}{x}{\chi}{\Gamma}{\sigma}$.
\begin{quote}
  To know $\FRMemEq{\Upsilon}{\Psi,x:\chi}{N_0}{N_1}{\phi}$ is to know, for any renaming
  \begin{tikzcd}[cramped]
    \IMode{\Upsilon} \arrow[r,hook,"\IMode{\rho}"] &\IMode{\Upsilon'}
  \end{tikzcd}
  and closed terms $\Member{M_0,M_1}{\Exprs_\sigma(\Upsilon')}$, that from
  $\FRMemEq{\Upsilon'}{\Psi\rho}{M_0}{M_1}{\chi\rho}$, you can conclude
  $\FRMemEq{\Upsilon'}{\Psi\rho}{\Subst{M_0}{x}{N_0\rho}}{\Subst{M_1}{x}{N_1\rho}}{\Subst{M_0}{x}{\phi\rho}}$.
  In other words:
  %
  \[
    \infer{
      \FRMemEq{\Upsilon}{\Psi,x:\psi}{N_0}{N_1}{\phi}
    }{
      \begin{array}{l}
        \forall
        \begin{tikzcd}[cramped, ampersand replacement = \&]
          \IMode{\Upsilon} \arrow[r,hook,"\IMode{\rho}"] \&\IMode{\Upsilon'}
        \end{tikzcd}.\
        \forall \Member{M_0,M_1}{\Exprs_\sigma(\Upsilon')}.\\
        \left(\FRMemEq{\Upsilon'}{\Psi\rho}{M_0}{M_1}{\chi\rho}\right)
        \Rightarrow
        \left(%
          \FRMemEq{\Upsilon'}{\Psi\rho}{%
            \Subst{M_0}{x}{N_0\rho}
          }{%
            \Subst{M_1}{x}{N_1\rho}%
          }{\Subst{M_0}{x}{\phi\rho}
          }%
        \right)
      \end{array}
    }
  \]
\end{quote}
%%
%% \section{Constructions on Refinements}
%%
%% Refinements can be arranged in a partial order via the
%% $\IsSubrefinement{\Phi}{\phi}{\psi}{\tau}$ judgment, and admit a lattice structure
%% at any sort $\tau$.
%%
%% We have the bottom refinement $\GloballyRefines{\TyBot{\tau}}{\tau}$, which simply
%% contains no elements. Next, the top refinement $\GloballyRefines{\TyTop{\tau}}{\tau}$
%% identifies all terms $M,N$ such that $\IsAbt{\cdot}{\Upsilon}{\cdot}{M}{\tau}$
%% and $\IsAbt{\cdot}{\Upsilon}{\cdot}{N}{\tau}$. The join
%% $\Refines{\Phi}{\phi\TyJoin{\tau}\psi}{\tau}$ of two refinements
%% $\Refines{\Phi}{\phi}{\tau}$ and $\Refines{\Phi}{\psi}{\tau}$ is defined as follows:
%% \begin{gather*}
%%   \infer{
%%     \RMemEq{\Phi}{M}{N}{\phi\TyJoin{\tau}\psi}
%%   }{
%%     \RMemEq{\Phi}{M}{N}{\phi}
%%   }\qquad
%%   \infer{
%%     \RMemEq{\Phi}{M}{N}{\phi\TyJoin{\tau}\psi}
%%   }{
%%     \RMemEq{\Phi}{M}{N}{\psi}
%%   }
%% \end{gather*}
%%
%% Finally the meet $\Refines{\Phi}{\phi\TyMeet{\tau}\psi}{\tau}$ of two refinements
%% $\Refines{\Phi}{\phi}{\tau}$ and $\Refines{\Phi}{\psi}{\tau}$ is defined
%% dually:
%% \[
%%   \infer{
%%     \RMemEq{\Phi}{M}{N}{\phi\TyMeet{\tau}\psi}
%%   }{
%%     \RMemEq{\Phi}{M}{N}{\phi} &
%%     \RMemEq{\Phi}{M}{N}{\psi}
%%   }
%% \]
%%
%% It is trivial to verify that these constructions on refinements give rise to a
%% lattice.
%%
%% \section{Multi-Sorted Computational Type Theory}
%%
%% Martin-L\"of's type theory is a theory of \emph{value types}: to define a type,
%% you specify how to form its equal canonical inhabitants, or values. Then, the
%% inhabitants of the types are explained uniformly by extending this relation
%% over an operational semantics such that two terms are equal members of a
%% type just when they compute to equal values of that type.
%%
%% More generally, one may wish to consider types which include computations in
%% addition to values; an example of such a type would be $\bar{\psi}$ when $\psi$
%% is a type, which would include all the members of $\psi$ as well as a divergent
%% computation. Another example is the type $\TyTop{\tau}$ which identifies all
%% closed terms of sort $\tau$.
%%
%% In order to ensure that such types are sensible, we must formulate a notion of
%% computational equivalence, and then say that a refinement is a \emph{type} just
%% when it respects computational equivalence; then, we are able to define types
%% as PERs on \emph{programs} in addition to values.
